---
title: "Beginner `climr` workflow"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
description: >
  Learn the basic `climr` workflow.
vignette: >
  %\VignetteIndexEntry{Beginner `climr` workflow}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
link-citations: true
code_folding: show
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.dim = c(7, 5),
  fig.align = "center"
)

if (!require(ggplot2)) {
  warning("'ggplot2' is not installed. Some visuals won't be available.")
}

cpath <- if (interactive()) {
  normalizePath("climr.cache", winslash = "/")
} else {
  normalizePath(file.path(tempdir(), "climr.cache"), winslash = "/")
}
options("climr.cache.path" = cpath)
```

### Using the `downscale()` function

The basic use of `climr` is to pass a `data.table` of point coordinates,
IDs and elevation to `downscale()` and select the type of climate
projection or values wanted - e.g. below we ask for future projections
in two periods, using two emissions scenarios and two General
Circulation Models. We also specify we want 3 individual runs of each
model/scenario combination, in addition to the ensemble mean (returned
by default).

```{r}
library(climr)

## provide a data.frame or data.table of point coordinates, IDs and elevation
my_points <- data.frame(
  lon = c(-123.4404, -123.5064, -124.2317),
  lat = c(48.52631, 48.46807, 49.21999),
  elev = c(52, 103, 357),
  id = seq_len(3)
)

## climr query for the data.frame
ds_out <- downscale(
  xyz = my_points,
  gcms = c("GFDL-ESM4", "EC-Earth3"), # specify two global climate models
  ssps = c("ssp370", "ssp245"), # specify two greenhouse gas concentration scenarios
  gcm_periods = c("2001_2020", "2041_2060"), # specify two 20-year periods
  max_run = 3, # specify 3 individual runs for each model
  vars = c("PPT", "CMD", "CMI")
)
```

The resulting output is a data.table, printed below:

```{r, echo = FALSE, results = "asis"}
knitr::kable(ds_out[id == head(id, 1)], caption = "'ds_out' table - Output from `downscale`. Outputs are shown for one location ('id')")
```

### Listing available options

The `list_*` functions (see ?list_gcm) are useful to see available options, for example:

```{r, eval = TRUE}
list_gcms()
list_gcm_periods()
list_ssps()
```

These functions can be used as arguments in `downscale()`. For example, `vars = list_variables()` could be specified to obtain data for all available variables, or `gcms = list_gcm()[c(1,4,5,6,7,10,11,12)]` could be used to specify the eight-gcm ensemble recommended by Mahony et al. (2022). 

### Working with raster data
A common use of `climr` is to obtain climate data for a user-provided raster grid. Here is some sample code showing how to do so. We convert a digital elevation model (DEM) into a table of points (one per grid cell) which we use as the `climr` input table. Then we take the `climr` input and populate the raster grid with a climate variable. 

```{r, eval = TRUE}
library(terra)
library(data.table)

## get the sample digital elevation model (dem) provided with `climr`
dem_vancouver <- get(data("dem_vancouver")) |> 
  unwrap()

## convert the DEM to a data.frame
my_grid <- as.data.frame(dem_vancouver, cells = TRUE, xy = TRUE)
colnames(my_grid) <- c("id", "lon", "lat", "elev") # rename column names to what climr expects

## A simple climr query. This will return the observed 1961-1990 and 2001-2020 mean annual temperature (MAT) for the raster grid points. 
ds_out <- downscale(
  xyz = my_grid, 
  obs_periods = "2001_2020", 
  vars = c("MAT")
)

## populate the raster grid with the downscaled climate values
my_clim <- dem_vancouver # use the DEM as a template raster
my_clim[ds_out[PERIOD == "2001_2020", id]] <- ds_out[PERIOD == "2001_2020", MAT] # populate the raster cells with the 2001-2020 mean annual temperature (MAT) values, using the `id` field as the link. 

plot(my_clim, main = "2001-2020 mean annual temperature (MAT)")

```

See `vignette("vignettes/climr_with_rasters.Rmd")` for more details on using `climr` with rasters and spatial data. 

### Looking under the hood

You can download and investigate the raw data that `climr` uses for downscaling, such as the reference climatological maps, the observational time series, and the GCM anomalies. See the section "Workflow with `*_input` functions and `downscale`" in
`vignette("vignettes/climr_workflow_int.Rmd")` for an example.
