---
title: "`climr` workflow - beginner"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
description: >
  Learn the basic `climr` workflow.
vignette: >
  %\VignetteIndexEntry{`climr` workflow - beginner}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
link-citations: true
code_folding: show
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.dim = c(7, 5),
  fig.align = "center"
)

if (!require(ggplot2)) {
  warning("'ggplot2' is not installed. Some visuals won't be available.")
}

cpath <- if (interactive()) {
  normalizePath("climr.cache", winslash = "/")
} else {
  normalizePath(file.path(tempdir(), "climr.cache"), winslash = "/")
}
options("climr.cache.path" = cpath)
```


The most basic use of `climr` is to pass a `data.table` of point coordinates, IDs
and elevation to `climr_downscale()` and select the type of climate projection or 
values wanted - e.g. below we ask for future projections in three periods, using
two emissions scenarios and two General Circulation Models. We also specify we want
3 individual runs of each model/scenario combination, in addition to the ensemble mean
(returned by default).

```{r}
library(climr)

## A data.table of point coordinates, IDs and elevation
data("xyzDT")

## if you just want to downscale points and not think about what happening behind the scenes, use this function
ds_out <- climr_downscale(
  xyz = xyzDT, which_normal = "auto",
  gcm_models = c("ACCESS-ESM1-5", "EC-Earth3"),
  ssp = c("ssp370", "ssp245"),
  gcm_period = c("2021_2040", "2041_2060", "2061_2080"),
  # gcm_ts_years = 2020:2060,
  max_run = 3, # we want 3 individual runs for each model
  vars = c("PPT", "CMD", "CMI")
)
```

These functions are useful to list available options:

```{r, eval = FALSE}
list_gcm()
list_gcm_period()
list_ssp()
list_variables()
```

```{r, echo = FALSE, results = "asis"}
knitr::kable(ds_out[id == head(id, 1)], caption = "'ds_out' table - Output from `climr_downscale`. Outputs are shown for one location ('id')")
```

Alternatively, you may want to download and investigate the normals and anomalies
before downscaling. See the section "Workflow with `*_input` functions and `downscale`"
in `vignette("vignettes/climr_workflow_int.Rmd")` for an example.

#### Adding / Modifying climate variables

All climate variables are computed via [R/append_clim_vars.R](./R/append_clim_vars.R).
Add or modify entries in the variables list as needed.
Use function `v` to handle dependencies (i.e. `v("Tmin")` instead of `dt$Tmin` / `dt[["Tmin"]]`).

#### Lapse rates details

See `vignette("vignettes/lapse_rates.Rmd")` for details.

#### Use of terra package

Instead of implementing custom algorithms for bilinear interpolation and raster manipulation, this package use `terra`.
Long term, it will benefit from any performance improvement in `terra`.
`terra` is a replacement for the `raster` package.

Data pivot using `data.table` dcast is the most resource expensive operation of the `downscale` function according to profiling using `profvis`.

