---
title: "Working with rasters and other spatial objects"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
description: >
  Learn how to use `climr` when working with spatial 
  input and output objects.
vignette: >
  %\VignetteIndexEntry{Working with rasters and other spatial objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
bibliography: references.bib
link-citations: true
code_folding: show
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.dim = c(7, 5),
  fig.align = "center"
)

options("climr.cache.path" = file.path("climr.cache"))
```

## Spatial inputs and `climr_downscale()`

We often start with a spatial object of a study area (in raster or
vector format) or point locations for which we want to obtain downscaled
climate variables.

Bellow we show how to work with these spatial objects and
`climr_downscale()`.

### Raster inputs

Often we have a raster of interest of the study area for which we want
full coverage of downscaled climate variables.

Since `climr_downscale()` accepts a table of point locations, we need to
convert the raster to a table of cell coordinates, with elevation
('elev') information and unique point IDs.

#### Raster with elevation values (a DEM)

If the raster is a DEM, the points `SpatVector` inherits the elevation
cell values.

The DEM we use below has 0 elevation values in ocean areas. Since
`climr` is meant to be used to downscale climate variables in land, we
will "clip" (set values outside the polygon to NAs) the raster using a
land-only polygon.

```{r, eval = FALSE, results = "hide"}
library(climr)
library(terra)
library(data.table)

dem_vancouver <- get(data("dem_vancouver")) |>
  unwrap()

vancouver_poly <- get(data("vancouver_poly")) |>
  unwrap()

dem_vancouver <- crop(dem_vancouver, vancouver_poly)
```

```{r, echo = FALSE, results = "hide", fig.cap = "Digital Elevation Model of North Vancouver w/ and w/o values outside polygon area", fig.show = "hold", out.width = "100%"}
library(climr)
library(terra)
library(data.table)

dem_vancouver <- get(data("dem_vancouver")) |>
  unwrap()

vancouver_poly <- get(data("vancouver_poly")) |>
  unwrap()

par(mfrow = c(1,2))

cols <- hcl.colors(100, "Earth")
plot(dem_vancouver, main = "Original DEM", col = cols)

## exclude areas outside the polygons
dem_vancouver <- mask(dem_vancouver, vancouver_poly)
plot(dem_vancouver, main = "Masked DEM", col = cols)

par(mfrow = c(1,1))
```

When making a table of raster cell (centroid) coordinates we must
preserve the cell IDs (with `as.data.frame(..., cell = TRUE)`) so that
we can later convert `clirm_downscale()` outputs back into raster
format.

```{r, results = "hide"}
## xy = TRUE ensures we get cell (centroid) coordinates
## cell = TRUE ensures we get cell ID numbers, which will be the point location IDs.
xyzDT <- dem_vancouver |>
  as.data.frame(., xy = TRUE, cell = TRUE) |>  
  as.data.table()

## rename colums to climr standard
setnames(xyzDT, c("id", "lon", "lat", "elev"))
```

Note that cells with NA's were automagically excluded.

```{r}
xyzDT
```

We can now pass our table of cell locations to `climr_downscale()`:

```{r, results = "hide"}
## in this case there's really no need for a spatial output format
dwscl_out <- climr_downscale(xyzDT, which_normal = "auto",
                             gcm_models = list_gcm()[1:3],
                             ssp = list_ssp()[1],
                             gcm_period = list_gcm_period(),
                             max_run = 0L, return_normal = FALSE,
                             vars = "MAT", out_spatial = FALSE)
```

To convert the outputs back into raster format, we simply need to use
the original raster use to extract the point locations as a template to
make rasters of downscaled climate values.

Since the location `id`s match raster cell IDs, the `dwscl_out$id`
column is used to index raster cells climate values when assigning
downscaled `MAT` values.

```{r}
tempRas <- rast(dem_vancouver)

dwscl_Ras <- sapply(split(dwscl_out, by = c("GCM", "PERIOD")) , function(DT, tempRas) {
  tempRas[DT$id] <- DT$MAT
  return(tempRas)
}, tempRas = tempRas)

## make a stack
dwscl_Ras <- rast(dwscl_Ras)

dwscl_Ras
```

```{r, echo = FALSE, results = "hide", fig.show = "hold", fig.cap = "`climr_downscale()` outputs converted to raster layers", fig.dim = c(15,9), out.width = "100%"}
minmax <- range(dwscl_out$MAT)

gcmstr <- paste(list_gcm()[1:3], collapse = "|")
titles <- sub(paste0("(", gcmstr, ")\\.([[:digit:]]{4})_([[:digit:]]{4})"), "\\1\\\n\\2-\\3", names(dwscl_Ras)) 
plot(dwscl_Ras, main = titles, nc = 5, nr = 3,
     type = "continuous", col = hcl.colors(100, "Reds", rev = 1), 
     range = minmax, plg = list(title = "MAT"), mar = c(0.1, 1, 1, 3))
```

#### Raster without elevation values

If the raster is not a DEM, we need to obtain elevation for each of the
points extracted from the raster cells. This can be done by extracting
elevation values from a DEM raster at the first raster's cell centroid
coordinates.

The DEM's resolution doesn't need to match the first raster's
resolution.

Here we start with a raster of biogeoclimatc zones for North Vancouver
and use a lower resolution DEM to obtain elevation values.

Again, we clip it to land areas.

```{r, results = "hide"}
## get point locations from raster of interest
BECz_vancouver_ras <- get(data("BECz_vancouver_ras")) |>
  unwrap()

BECz_vancouver_ras <- mask(BECz_vancouver_ras, vancouver_poly)

## raster to data.table and rename columns to climr standard
xyzDT <- BECz_vancouver_ras |>
  as.data.frame(., xy = TRUE, cell = TRUE) |>  
  as.data.table()
setnames(xyzDT, c("id", "lon", "lat", "ZONE"))

## now extract elevation values at each cell centroid coordinate
## from a DEM raster
dem_vancouver_lowres <- get(data(dem_vancouver_lowres)) |>
  unwrap()

xyzDT$elev <- extract(dem_vancouver_lowres, xyzDT[, .(lon, lat)])[names(dem_vancouver_lowres)]
```

```{r, fig.cap = "Rasters used to derive point locations (left) and elevation values (right)."}
par(mfrow = c(1,2))
plot(BECz_vancouver_ras, main = "BGC zones N Vancouver")
plot(dem_vancouver_lowres, main = "Low res. DEM",
     col = hcl.colors(100, "Earth"))
par(mfrow = c(1,1))
```

The process to convert the downscaled data to raster format is the same
as before, because we again used the raster cell IDs as point location
IDs (`xyzDT$id`):

```{r}
dwscl_out <- climr_downscale(xyzDT, which_normal = "auto",
                             gcm_models = list_gcm()[1:3],
                             ssp = list_ssp()[1],
                             gcm_period = list_gcm_period(),
                             max_run = 0L, return_normal = FALSE,
                             vars = "MAT", out_spatial = FALSE)

tempRas <- rast(BECz_vancouver_ras)

dwscl_Ras <- sapply(split(dwscl_out, by = c("GCM", "PERIOD")) , function(DT, tempRas) {
  tempRas[DT$id] <- DT$MAT
  return(tempRas)
}, tempRas = tempRas)

## make a stack
dwscl_Ras <- rast(dwscl_Ras)
```

#### Lowering an input raster resolution

We may want to downscale climate at a coarser resolution than the, e.g.,
DEM raster layer we have at hand:

```{r, results = "hide"}
vancouver_poly <- get(data("vancouver_poly")) |>
  unwrap()
dem_vancouver <- get(data("dem_vancouver")) |>
  unwrap()

dem_vancouver <- mask(dem_vancouver, vancouver_poly)

## extract points at a coarser scale, but elevation at original scale
lowRes_ras <- rast(dem_vancouver) 
lowRes_ras <- project(lowRes_ras, crs(lowRes_ras), res = 0.01) 

lowRes_ras[] <- 1
lowRes_ras <- mask(lowRes_ras, vancouver_poly) ## set water back to NA

lowRes_points <- as.points(lowRes_ras)
lowRes_points$id <- 1:nrow(lowRes_points)

xyzDT <- extract(dem_vancouver, lowRes_points, xy = TRUE) |>
  as.data.table()
setnames(xyzDT, c("id", "elev", "lon", "lat"))

## exclude NAs
xyzDT <- xyzDT[complete.cases(xyzDT)]
lowRes_points <- lowRes_points[lowRes_points$id %in% xyzDT$id,]
```

```{r, echo = FALSE, results = "hide", fig.cap = "Points extracted from raster", fig.show = "hold", out.width = "100%"}
plot(dem_vancouver, col = cols)
points(lowRes_points, alpha = 0.7)
```

As before, pass your `data.table` of point locations to
`climr_downscale`:

```{r, results = "hide"}
dwscl_out <- climr_downscale(xyzDT, which_normal = "auto",
                                   gcm_models = list_gcm()[1:3],
                                   ssp = list_ssp()[1],
                                   gcm_period = list_gcm_period(),
                                   max_run = 0L, return_normal = FALSE,
                                   vars = "MAT", out_spatial = TRUE)
```

```{r, echo = FALSE, results = "hide", fig.show = "hold", fig.cap = "Plot of spatial outputs from `climr_downscale()` using more spaced point locations", fig.dim = c(15,9), out.width = "100%"}
minmax <- range(dwscl_out$MAT)

dwscl_outLong <- as.data.table(dwscl_out, geom = "XY") |>
  dcast(formula = ... ~ GCM + PERIOD, value.var = "MAT") |>
  vect(geom = c("x", "y"), crs = crs(dem_vancouver))

lyrs <- setdiff(names(dwscl_outLong), c("elev", "id", "SSP", "RUN"))
gcmstr <- paste(list_gcm()[1:3], collapse = "|")
titles <- sub(paste0("(", gcmstr, ")_([[:digit:]]{4})_([[:digit:]]{4})"), "\\1\\\n\\2-\\3", lyrs) 
plot(dwscl_outLong, lyrs, main = titles, nc = 5, nr = 3,
     type = "continuous", col = hcl.colors(100, "Reds", rev = 1), 
     range = minmax, plg = list(title = "MAT"), mar = c(0.1, 1, 1, 3))
```

### Vector inputs/outputs

#### Point vector data

The process is again similar to the above.

If the points have elevation information we simply need to convert them
to a `data.table`. In this case, point IDs need to be added to the
`data.table` afterwards.

If they don't have elevation information, then we need to obtain it. In
the example below we start with a set of points in North Vancouver, BC,
Canada, for which we get elevation values from a Digital Elevation Model
(DEM) raster.

The `terra::extract()` function outputs a `data.frame` with point IDs,
the extracted values and, optionally x/y coordinates. So we need simply
to use this output to create our `xyzDT` table for `climr_downscale()`.

To get point-type outputs simply set
`climr_downscale(..., out_spatial = TRUE)`.

```{r, results = "hide"}
vancouver_points <- get(data("vancouver_points")) |>
  unwrap()
dem_vancouver <- get(data("dem_vancouver")) |>
  unwrap()

xyzDT <- extract(dem_vancouver, vancouver_points, xy = TRUE) |>
  as.data.table()
setnames(xyzDT, c("id", "elev", "lon", "lat"))

dwscl_out <- climr_downscale(xyzDT, which_normal = "auto",
                             gcm_models = list_gcm()[1:3],
                             ssp = list_ssp()[1],
                             gcm_period = list_gcm_period(),
                             vars = "MAT", max_run = 0L, 
                             return_normal = FALSE,
                             out_spatial = TRUE)
dwscl_out
```

```{r, echo = FALSE}
dwscl_out
```

```{r, echo = FALSE, results = "hide", fig.show = "hold", fig.cap = "Plot of spatial outputs from `climr_downscale()`", fig.dim = c(15,9), out.width = "100%"}
vancouver <- get(data("vancouver")) |>
  unwrap()
vancouver <- crop(vancouver, ext(vancouver_points))

minmax <- range(dwscl_out$MAT)

dwscl_outLong <- as.data.table(dwscl_out, geom = "XY") |>
  dcast(formula = ... ~ GCM + PERIOD, value.var = "MAT") |>
  vect(geom = c("x", "y"), crs = crs(dem_vancouver))

lyrs <- setdiff(names(dwscl_outLong), c("elev", "id", "SSP", "RUN"))
gcmstr <- paste(list_gcm()[1:3], collapse = "|")
titles <- sub(paste0("(", gcmstr, ")_([[:digit:]]{4})_([[:digit:]]{4})"), "\\1\\\n\\2-\\3", lyrs) 
plot(dwscl_outLong, lyrs, main = titles, nc = 5, nr = 3,
     type = "continuous", col = hcl.colors(100, "Reds", rev = 1), 
     range = minmax, plg = list(title = "MAT"), mar = c(0.1, 1, 1, 3))
```

### Polygon vector data

We assume here that your polygon data refers to an area (or several) for
which you want downscaled climate variables for.

Because each polygon is associated with a single value per attribute or
field, the polygons will need to be sampled/combined with the points
where downscaling will happen. In the examples above we showed how to
use a polygon to mask a raster or points dataset before creating the
`xyzDT` table, which would be a way to achieve this.

Here, we show another example of how points derived from the polygon
centroids.

#### using polygon centroids

**We advise *against* doing this unless your polygons are so small that
the centroid reflects the overall polygon's climate conditions.**

```{r, results = "hide"}
vancouver_poly <- get(data("vancouver_poly")) |>
  unwrap()
dem_vancouver <- get(data("dem_vancouver")) |>
  unwrap()

centrds <- centroids(vancouver_poly, inside = TRUE)

## get polygon areas and exclude very large polygon centroids
centrds$area <- expanse(vancouver_poly)
centrds <- centrds[centrds$area < 50000,]

## we retained only 3 points

xyzDT <- extract(dem_vancouver, centrds, xy = TRUE) |>
  as.data.table()
setnames(xyzDT, c("id", "elev", "lon", "lat"))

dwscl_outGrid <- climr_downscale(xyzDT, which_normal = "auto",
                                 gcm_models = list_gcm()[1:3],
                                 ssp = list_ssp()[1],
                                 gcm_period = list_gcm_period(),
                                 max_run = 0L, return_normal = FALSE,
                                 vars = "MAT", out_spatial = TRUE)
```

```{r, echo = FALSE, fig.show = "hold", fig.cap = "Excluded (grey) and retained (orange) polygon centroids", fig.dim = c(5,3)}
plot(vancouver_poly)
points(centroids(vancouver_poly), col = "grey")
points(centrds, col = "salmon1")
```

## References
